### 2017-06-13 01-intro.tex
Process Henri's comments 1june (except Maxwell/Pascal analysis for GPU strategies)

\section{Introduction}
The tremendous amount of data that we generate through our daily applications such as social networking services, online shopping, and news recommendations, provides us with an opportunity to extract hidden but useful, even invaluable information. Realizing this opportunity, however, requires a significant amount of effort because traditional machine learning algorithms often become extremely inefficient with large amounts of data.

There have been two main approaches to resolving this issue; machine learning researchers have developed new scalable algorithms~\cite{bottou2010large, boyd2011distributed}, while systems and networking researchers have worked on developing new generic infrastructure systems which can be leveraged to construct machine learning solutions more efficiently~\cite{dean2008mapreduce, chang2008bigtable}.
However, the best possible performance is often achieved by carefully integrating both of these approaches in a single solution.

@@

In this paper, we focus on the SG-MCMC method applied to the a-MMSB model. Recent
work of the authors of this paper~\cite{LiAW15} showed that this methodology is faster and more accurate than the
SVB method. That work further proposes a heuristic to scale up the number of
communities at the cost of less precision; the work in this paper considers the
algorithm without that heuristic.

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms like deep learning in several
ways. First, it is \emph{highly data-intensive} which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are non-deterministic. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
community graphs.

@@

\begin{itemize}
\item \textbf{scaling up to run on GPUs.}
The application is not a natural fit for GPUs, because it is data-bound and
irregular in its data accesses. Nevertheless, we developed aggressive
memory allocation optimizations to solve the data bandwidth issues for GPUs.
This implementation achieves up to 120$\times$ speedup compared to the baseline.
%
We managed to reduce the algorithm's memory requirements by roughly 75\%, which
significantly reduces the data intensity and allows for tackling larger
problems; % while maintaining all state in memory.

\item \textbf{scaling out to overcome the memory bounds posed by GPUs} or,
generally, single-machine systems, to solve the largest publicly available
community graphs. % This necessitated overcoming several challenges.
% The algorithm's state grows rapidly with larger graphs and number of latent
% communities.
Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned across a cluster
of machines. Hence, the cluster nodes must read
remote memory hosted by their peers. Since the algorithm is data-bound, we
substantially optimized remote memory accesses by implementing
a custom RDMA store.
On top of that, we pipelined the algorithm stages by fetching
data in advance over the network.
\begin{comment} Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.
\end{comment}

\begin{comment}
\item \textbf{the use of pipelining} \emph{in the distributed implementation}
justifies straightforward intra-node thread parallelism, and it is pointless
to combine the GPU implementations
into it. The cause is that computing for an iteration takes less
time than fetching the data from remote memory, and the computation is
fully overlapped with fulfilling the data requests for the next iteration.
Hence, it is useless to improve the computation time.
\end{comment}

@@

A bird's eye-view of the work in this paper, without any technical detail, has been
published at CCGrid-2016~\cite{10.1109/CCGrid.2016.98}.
The results of our distributed MCMC-aMMSB implementation
have been previously published at the 2016 ParLearning
workshop~\cite{DBLP:conf/ipps/El-HelwHLAWB16}, where it earned a best-paper
award.

>>>>

\section{Introduction}
The tremendous amount of data that we generate through our daily applications such as
social networking services, online shopping, and news recommendations, provides us
with an opportunity to extract hidden but useful, even invaluable information.
Realizing this opportunity, however, requires a significant amount of effort because
traditional machine learning algorithms may become inefficient with large amounts of data.

There have been two main approaches to resolving this issue; machine learning researchers have developed new scalable algorithms~\cite{bottou2010large, boyd2011distributed}, while systems and networking researchers have worked on developing new generic software infrastructure systems which can be leveraged to construct machine learning solutions more efficiently~\cite{dean2008mapreduce, chang2008bigtable}.
However, the best possible performance is often achieved by carefully integrating both of these approaches in a single solution.

@@

In this paper, we focus on the SG-MCMC method applied to the a-MMSB model. Recent
work of the authors of this paper~\cite{LiAW15} showed that this methodology is faster and more accurate than the
SVB method.

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms like deep learning in several
ways. First, it is \emph{highly data-intensive} which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are \emph{non-deterministic}. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
community graphs.

@@

\begin{itemize}
\item \textbf{scaling up to run on GPUs.}
The application is not a natural fit for GPUs, because it is data-bound and
irregular in its data accesses. Nevertheless, we developed aggressive
memory allocation optimizations to solve the data bandwidth issues for GPUs.
This implementation achieves up to 120$\times$ speedup compared to the C++ baseline.
%
We reduced the algorithm's memory requirements by roughly 75\%, which
significantly reduces the data intensity and allows for tackling larger
problems; % while maintaining all state in memory.

\item \textbf{scaling out to overcome the memory bounds} posed by 
single-machine systems, to be the first to solve the largest publicly available
community graphs. % This necessitated overcoming several challenges.
% The algorithm's state grows rapidly with larger graphs and number of latent
% communities.
Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned across a cluster
of machines. Hence, the cluster nodes must read
remote memory hosted by their peers. Since the algorithm is data-bound, we
substantially optimized remote memory accesses by implementing
a custom RDMA store.
On top of that, we pipelined the algorithm stages by fetching
data in advance over the network.
\begin{comment} Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.
\end{comment}

The use of pipelining makes it pointless to combine GPU acceleration
into the distributed implementation: computing for an iteration takes less
time than fetching the data from remote memory, and the computation is
fully overlapped with fulfilling the data requests for the next iteration.
% Hence, it is useless to improve the computation time.

@@

A bird's-eye view of the work in this paper, without any technical detail, has been
published at CCGrid-2016~\cite{10.1109/CCGrid.2016.98}.
The results of our distributed MCMC-aMMSB implementation
have been previously published at the 2016 ParLearning
workshop~\cite{DBLP:conf/ipps/El-HelwHLAWB16}, where it earned a best-paper
award.

### 2017-06-13 06-conclusion.tex
Process Henri's comments 1june (except Maxwell/Pascal analysis for GPU strategies)

In our implementation for accelerators, we navigated the complex optimization
landscape by dynamically generating kernel code for finest control of allocation in
the GPU memory hierarchy, and by tuning parameters like kernel block size and vector
width. These efforts resulted in
a significant speedup of 21$\times$ on a GPU.
This speedup is achieved in comparison to an already
optimized sequential implementation. The evaluation of the performance
across several GPUs yielded one striking find. No single optimization strategy
works best for all GPUs: different choices for memory allocation strategy,
kernel block size and vector width were required for optimal results for
different GPUs and problem instances. Counter-intuitively, some GPUs performed
best with vectorization disabled.
%
The outcome of this study therefore reinforces the significance of avoiding premature
optimization. In particular, the success
of common GPU optimizations depends on the device and the
problem size it is applied to.

We further created a highly scalable implementation to handle the largest existing
community graphs on a distributed cluster machine.
We made several contributions to
achieve scalability and high performance:
the most important data structures are partitioned across the collective memory of the
cluster nodes;
we use a
mixture of MPI and RDMA primitives to speed up the communication between cluster
nodes; and
we overlapped computation
with communication to hide latency.
%
By evaluating the system empirically using 65 cluster nodes and large data sets,
we show that its strong and weak scalability are both good. We also compared its
performance against a 40-core shared-memory system with 1TB of memory.
The distributed system has far stronger performance because it has more cores and
more memory bandwidth, although it must fetch its data through the network.
% Additionally, we assessed the efficiency of the algorithm's resource utilization.
Finally, a
demonstration of the implementation's utility was provided by processing 6
large real-world data sets.
To the best of
our knowledge, this is the first time that the problem of deducing overlapping
communities has been learned for problems of such a large scale.

>>>>

In our implementation for accelerators, we navigated the complex optimization
landscape by dynamically generating kernel code for finest control of allocation in
the GPU memory hierarchy, and by tuning parameters like kernel block size and vector
width. These efforts resulted in
a significant speedup of 120$\times$ on a GPU.
This speedup is achieved in comparison to an already
optimized sequential implementation. The evaluation of the performance
across several GPUs yielded one striking find. No single optimization strategy
works best for all GPUs: different choices for memory allocation strategy,
kernel block size and vector width were required for optimal results for
different GPUs and problem instances. Counter-intuitively, some GPUs performed
best with vectorization disabled.
%
The outcome of this study therefore reinforces the significance of avoiding premature
optimization. In particular, the success
of common GPU optimizations depends on the device and the
problem size it is applied to.

We further created a highly scalable implementation to handle the largest existing
community graphs on a distributed cluster machine. The most important data structures
are partitioned across the collective memory of the cluster nodes, which
significantly increases data access latency for our already data-dominant algorithm.
We made several contributions to improve performance and still maintain scalabitiy:
we created a custom distributed Key-Value store implemented with high-speed RDMA
primitives that exploits the algorithm's structure to avoid concurrency issues; we
use MPI for synchronization and work distribution between the master and worker nodes;
and
we overlapped computation
with communication to hide latency.
%
By evaluating the system empirically using 65 cluster nodes and large data sets,
we show that its strong and weak scalability are both good. We also compared its
performance against a 40-core shared-memory system with 1TB of memory.
The distributed system has far stronger performance because it has more cores and
more collective memory bandwidth, even though it must fetch its data through the network.
% Additionally, we assessed the efficiency of the algorithm's resource utilization.
Finally, a
demonstration of the implementation's utility was provided by processing 6
large real-world data sets.
To the best of
our knowledge, this is the first time that the problem of deducing overlapping
communities has been learned for problems of such a large scale.

### 2017-06-13 distr-03-design.tex
Process Henri's comments 1june (except Maxwell/Pascal analysis for GPU strategies)

This section describes how we create and evaluate a distributed version of the C++
algorithm. We used the multicore CPUs in the DAS5 with straightforward
intra-node thread parallelisation, not the GPU implementations described
in the previous section. The justification is twofold: in our DAS5 system,
it is pointless to speed up the intra-node performance beyond straightforward
multithreading, because distributed performance is dominated by remote memory
latency; the analysis in Section~\ref{eval-pipeline} shows that the time for
(pre)loading the data from remote hosts greatly exceeds the calculation time.
The calculation time is fully overlapped in our pipelining implementation.
Moreover, usage of GPUs in the distributed setting would necessitate staging of
the relevant parts of $\pi/\phi$ from the host to GPU memory in each iteration,
which would eat most of the performance gain offered by the GPU because
the algorithm is so data-intensive.
% The details of the multi-threaded parallelism
% will be discussed only where appropriate.
The distributed design
follows a master-worker paradigm, where the master controls the sequence of
parallel operations and the workers perform the calculations. For thread
parallelism, we annotate the program with OpenMP~\cite{OpenMPSpec}. For
distributed communication, we use MPI~\cite{Forum:1994:MMI:898758}.

>>>>

This section describes how we create and evaluate a distributed version of the C++
algorithm. We used multicore CPUs with straightforward
intra-node thread parallelisation, not the GPU implementations described
in the previous section. The justification is twofold:
it is pointless to speed up the intra-node performance beyond straightforward
multithreading, because distributed performance is dominated by remote memory
latency; the analysis in Section~\ref{eval-pipeline} shows that the time for
(pre)loading the data from remote hosts greatly exceeds the calculation time.
The calculation time is fully overlapped in our pipelining implementation.
Secondly, usage of GPUs in the distributed setting would necessitate staging of
the relevant parts of $\pi/\phi$ from the host to GPU memory in each iteration,
which would consume most of the performance gain offered by the GPU because
the algorithm is so data-intensive.
% The details of the multi-threaded parallelism
% will be discussed only where appropriate.

The distributed design
follows a master-worker paradigm, where the master controls the sequence of
parallel operations and the workers perform the calculations. For thread
parallelism, we annotate the program with OpenMP~\cite{OpenMPSpec}. For
distributed communication, we use MPI~\cite{Forum:1994:MMI:898758}.

### 2017-06-13 gpu-05-03-accelerator.tex
NVidia iso Nvidia, small rephrase
Process Henri's comments 1june (except Maxwell/Pascal analysis for GPU strategies)

\subsubsection{Analysis of GPU Parallelism}
As discussed in Section~\ref{gpu-design}, there are 10 distinct flavors of the
GPU version of the algorithm. Specifically, there is the naive strategy, the
shared strategy and 8 variations of the code generation strategy. This
section investigates the effectiveness of each flavor on Nvidia's GTX~\mbox{Titan-X}
Maxwell and Pascal,
GTX980, K40c and K20m.

% TITANX
\begin{figure*}[tb]	% [htb]
  \centering
  \epsfig{file=plots/gpu-sweep-legends.eps, width=1.0\textwidth}
  \setlength\tabcolsep{0em}
  \begin{tabular}{c c c}
  \epsfig{file=plots/titanx-1024-strategies-w1.eps, width=0.33\textwidth} &
  \epsfig{file=plots/titanx-2048-strategies-w1.eps, width=0.33\textwidth} &
  \epsfig{file=plots/k40-1024-strategies-w2.eps, width=0.33\textwidth}
  \end{tabular}
  \caption{Execution time of 1000 \textit{update\_phi} invocations across a sweep of
  	\textit{update\_phi} thread block sizes. The lower figures are a zoom-in into
	the optimal block range of the upper figures.
	Other relevant model parameters: M=4096, $|\Neighbors|$=32.
	% (a) \mbox{Titan-X} GPU, without explicit kernel vectorization;
	% Relevant model parameters: K=1024, M=4096, $|\Neighbors|$=32.
  	% (b) \mbox{Titan-X} GPU, without explicit kernel vectorization;
	% K=2048, M=4096, $|\Neighbors|$=32.
	% (c) K40c GPU, with explicit kernel vectorization of width 2;
	% K=1024, M=4096, $|\Neighbors|$=32.
  }
  % \label{titanx-w1-sweep}
  \label{gpu-sweep}
\end{figure*}

\begin{comment}
% TITANX K=2K
\begin{figure*}[tb]	% [htb]
  \centering
  \epsfig{file=plots/titanx-2048-strategies-w1.eps, width=\textwidth}
  \caption{Execution time of 1000 \textit{update\_phi} invocations using the
  \mbox{Titan-X(Maxwell)} GPU,
  without explicit kernel vectorization, across a sweep of
  \textit{update\_phi} thread block
  sizes. Relevant model parameters: K=2048, M=4096, $|\Neighbors|$=32.}
  \label{titanx-w1-sweep-2k}
\end{figure*}
\end{comment}

Figure \ref{gpu-sweep}(a) presents the performance of the Maxwell \mbox{Titan-X} GPU
for all 10 strategy flavors, without explicitly vectorizing the kernels. The
x-axis represents different \textit{update\_phi} thread block sizes while the y-axis presents
the total execution time of 1000 invocations of the \textit{update\_phi} kernel. The
naive and shared strategies are labeled NAIVE and SHARED respectively. Further,
each flavor of the code generation strategy is labeled by GEN followed by the 3
choices that identify it.
%
The wide performance range inhibits the readability of the figure,
therefore, a more focused Figure~\ref{gpu-sweep}(a), bottom figure, is provided. As would
be expected, the naive strategy exhibits the worst performance over all
thread block
sizes, as it does not explicitly cache repeated device memory read operations.
The SHARED and \textit{GEN-SSS} strategies come next in terms of performance.
Both strategies cache $Grads_i$, $Probs_i$ and $\pi_i$ in shared memory but
differ in one aspect. Namely, \textit{GEN-SSS} explicitly unrolls the internal
loops of the kernel. However, there is no significant performance difference
between them. The other flavors of the code generation strategy attain higher
performance as they unroll internal loops as well as cache data in
registers. The Pascal Titan-X is fastest with the \textit{GEN-RSR} strategy and a
block size of 96 and a vector width of~2, whereas
the Maxwell Titan-X obtains the best performance with the \textit{GEN-RSS}
strategy and a thread block size of 64 and no vectorization. The results of other
vector widths
are omitted as they obtain lower performance.

A key model parameter that affects the behavior of the optimization strategies
is the number of communities $K$. Figure~\ref{gpu-sweep}(b) presents the
same model configuration as in Figure~\ref{gpu-sweep}(a) but $K=2048$
instead of $K=1024$. One key difference between the two figures is the optimal
thread block size. An increase in $K$ comes with a proportional increase in
the size of shared memory space that is used by each thread block for the
strategies that employ shared memory. Similarly, GEN strategies that use the
register file will require additional space. Therefore, the number of
concurrent thread blocks that can execute on a single streaming multiprocessor
will decrease, minimizing the GPU's occupancy and utilization. This effect is
most clear when comparing the NAIVE with the SHARED and GEN-SSS strategies. In
this case, the NAIVE strategy outperforms both SHARED and GEN-SSS for a
thread block
size of 32 due to their low occupancy. This limitation can be counteracted by
selecting a larger thread block size which in turn increases the computation
concurrency and occupancy.

@@

\subsubsection{Comparison of Compute Devices}
Figure~\ref{gpus-fig} compares the highest speedup achieved by each of the
GTX~\mbox{Titan-X} Maxwell and Pascal, GTX980, K40c and K20m relative to the baseline
sequential C++ version. These results are consistent with the relative
capabilities of each device as listed in Table~\ref{table-gpus}. For instance,
the \mbox{Titan-X(Pascal)} achieves the highest speedup of 120 relative to the baseline.

>>>>

\subsubsection{Analysis of GPU Parallelism}
As discussed in Section~\ref{gpu-design}, there are 10 distinct flavors of the
GPU version of the algorithm. Specifically, there is the naive strategy, the
shared strategy and 8 variations of the code generation strategy. This
section investigates the effectiveness of each flavor on NVidia's GTX~\mbox{Titan-X}
Maxwell and Pascal,
GTX980, K40c and K20m.

% TITANX
\begin{figure*}[tb]	% [htb]
  \centering
  \epsfig{file=plots/gpu-sweep-legends.eps, width=1.0\textwidth}
  \setlength\tabcolsep{0em}
  \begin{tabular}{c c c c}
  \epsfig{file=plots/titanx-maxwell-1024-strategies-w2.eps, width=0.25\textwidth} &
  \epsfig{file=plots/titanx-maxwell-2048-strategies-w2.eps, width=0.25\textwidth} &
  \epsfig{file=plots/titanx-pascal-1024-strategies-w2.eps, width=0.25\textwidth} &
  \epsfig{file=plots/k40-1024-strategies-w2.eps, width=0.25\textwidth}
  \end{tabular}
  \caption{Execution time of 1000 \textit{update\_phi} invocations across a sweep of
  	\textit{update\_phi} thread block sizes. The lower figures are a zoom-in into
	the optimal block range of the upper figures.
	Other relevant model parameters: M=4096, $|\Neighbors|$=32.
	% (a) \mbox{Titan-X(Pascal)} GPU, with kernel vectorization of width 2;
	% Relevant model parameters: K=1024, M=4096, $|\Neighbors|$=32.
	% (b) \mbox{Titan-X} GPU, without explicit kernel vectorization;
	% Relevant model parameters: K=1024, M=4096, $|\Neighbors|$=32.
  	% (c) \mbox{Titan-X} GPU, without explicit kernel vectorization;
	% K=2048, M=4096, $|\Neighbors|$=32.
	% (d) K40c GPU, with explicit kernel vectorization of width 2;
	% K=1024, M=4096, $|\Neighbors|$=32.
  }
  % \label{titanx-w1-sweep}
  \label{gpu-sweep}
\end{figure*}

\begin{comment}
% TITANX K=2K
\begin{figure*}[tb]	% [htb]
  \centering
  \epsfig{file=plots/titanx-maxwell-2048-strategies-w1.eps, width=\textwidth}
  \caption{Execution time of 1000 \textit{update\_phi} invocations using the
  \mbox{Titan-X(Maxwell)} GPU,
  without explicit kernel vectorization, across a sweep of
  \textit{update\_phi} thread block
  sizes. Relevant model parameters: K=2048, M=4096, $|\Neighbors|$=32.}
  \label{titanx-w1-sweep-2k}
\end{figure*}
\end{comment}

Figure \ref{gpu-sweep}(a) presents the performance of the Pascal \mbox{Titan-X} GPU
for all 10 strategy flavors, with an explicit vector width~2 in the kernels. The
x-axis represents different \textit{update\_phi} thread block sizes while the y-axis presents
the total execution time of 1000 invocations of the \textit{update\_phi} kernel. The
naive and shared strategies are labeled NAIVE and SHARED respectively. Further,
each flavor of the code generation strategy is labeled by GEN followed by the 3
choices that identify it.
%
The wide performance range inhibits the readability of the figure,
therefore, a zoomed-in version of Figure~\ref{gpu-sweep}(a), is provided as the bottom figure. As would
be expected, the naive strategy exhibits the worst performance over all
thread block
sizes, as it does not explicitly cache repeated device memory read operations.
The SHARED and \textit{GEN-SSS} strategies come next in terms of performance.
Both strategies cache $Grads_i$, $Probs_i$ and $\pi_i$ in shared memory but
differ in one aspect: \textit{GEN-SSS} explicitly unrolls the internal
loops of the kernel. However, there is no significant performance difference
between them. The other flavors of the code generation strategy attain higher
performance as they unroll internal loops as well as cache data in
registers. The Pascal Titan-X is fastest with the \textit{GEN-RSR} strategy and a
block size of 96 and a vector width of~2, whereas
the Maxwell Titan-X in Figure~\ref{gpu-sweep}(b) obtains the best performance with the \textit{GEN-RSS}
strategy and a thread block size of 64 and no vectorization. The results of other
vector widths
are omitted as they obtain lower performance.

A key model parameter that affects the behavior of the optimization strategies
is the number of communities~$K$. Figure~\ref{gpu-sweep}(c) presents the
same model configuration as in Figure~\ref{gpu-sweep}(b) but $K=2048$
instead of $K=1024$. The most important difference in performance between the two figures is the optimal
thread block size. An increase in $K$ comes with a proportional increase in
the size of shared memory space that is used by each thread block for the
strategies that employ shared memory. Similarly, GEN strategies that use the
register file will require additional space. Therefore, the number of
concurrent thread blocks that can execute on a single streaming multiprocessor
will decrease, minimizing the GPU's occupancy and utilization. This effect is
most clear when comparing the NAIVE with the SHARED and GEN-SSS strategies. In
this case, the NAIVE strategy outperforms both SHARED and GEN-SSS for a
thread block
size of 32 due to their low occupancy. This limitation can be counteracted by
selecting a larger thread block size which in turn increases the computation
concurrency and occupancy.

@@

\subsubsection{Comparison of Compute Devices}
Figure~\ref{gpus-fig} compares the highest speedup achieved by each of the
GTX~\mbox{Titan-X} Maxwell and Pascal, GTX980, K40c and K20m relative to the baseline
sequential C++ version. These results are consistent with the relative
capabilities of each device as listed in Table~\ref{table-gpus}. For instance,
the Pascal \mbox{Titan-X} achieves the highest speedup of 120 relative to the baseline.
